---
title: "LSE stock-price regression and variable selection"
author: "leon5354"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    toc: yes
    toc_depth: 2
urlcolor: blue
---

```{r setup, include=FALSE}
project_root <- rprojroot::find_root(rprojroot::has_file("README.md"))
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

## Introduction

This report builds an explanatory regression model for **VOD** (Vodafone's
share price) using a basket of other London Stock Exchange–listed stocks as
predictors. The focus is not on forecasting — there is no train/test split and
no time-series modelling — but on the *variable-selection* workflow:

1. Screen predictors for multicollinearity.
2. Apply four selection strategies (leaps-and-bounds, stepwise, forward,
   backward).
3. Use Box-Cox transforms on the response to linearise residuals.
4. Check assumptions with the standard `lm` diagnostic plots and the residual
   ACF.

This is a refreshed version of a 2023 coursework project. The methodology is
unchanged; the dataset has been replaced (see Data below).

\bigskip
\noindent\textbf{Important — synthetic data.} The original 2023 coursework used
an LSE basket supplied with the assignment (`lse.RData`), which was never
committed and is no longer available. `data/generate_demo_data.R` produces a
synthetic `lse` data frame with the **same schema** (same column names) so the
analysis code runs end-to-end. The patterns the analysis finds
(multicollinearity structure, the VOD signal, the need for a response
transform) are reproduced qualitatively, but specific numbers — adj-$R^2$,
Box-Cox $\lambda$, coefficient estimates — are illustrative, not real.

## Data

The synthetic `lse` frame is generated by `data/generate_demo_data.R`. It
contains 1,304 daily observations of 27 stocks plus `Date`, `Weekday`,
`Year`, and `Month` columns. Prices follow correlated geometric Brownian
motion with a two-factor structure that induces the kind of multicollinearity
the selection workflow is designed to address.

```{r load-data}
require(car); require(ggplot2); require(dplyr); require(broom); require(ggpubr)
require(GGally); require(corrplot); require(leaps); require(MASS)

load("data/lse.RData")

lse1   <- subset(lse, select = -c(Date, Weekday))
cor1   <- cor(lse1)
```

```{r summary-stats}
data_stats <- data.frame(
  Statistic = c("Rows", "Numeric columns", "Missing values", "Date range"),
  Value = c(nrow(lse),
            sum(sapply(lse, is.numeric)) + 4,
            sum(is.na(lse)),
            paste(range(lse$Date), collapse = " to ")))
knitr::kable(data_stats, col.names = c("Statistic", "Value"),
             caption = "Dataset summary")
```

## Multicollinearity screening

The correlation heatmap reveals several clusters of highly correlated
predictors — typical of stock baskets, where most names co-move with the
market.

```{r corrplot, fig.height=6, fig.width=6.5, fig.cap="Correlation matrix of all numeric variables. Off-diagonal blocks reveal clusters of collinear predictors."}
corrplot::corrplot.mixed(cor1, lower.col = "black", number.cex = .4)
```

Following the original analysis, we drop the predictors that are most
collinear with the others and least correlated with VOD: `EXPN`, `SVT`,
`SMT`, and `Year` (kept only for grouping, not as a signal).

```{r prune}
lsenew <- subset(lse1, select = -c(EXPN, SVT, SMT, Year))
```

## Transformations

Scatterplots of VOD against the candidate predictors suggest log transforms
for `SSE`, `LLOY`, and `ABF` to tighten the relationship. Each is shifted to
be strictly positive before the log is applied.

```{r transforms, echo=TRUE}
lsenew$SSE.log  <- log10(lsenew$SSE  + 1 - min(lsenew$SSE))
lsenew$LLOY.log <- log(lsenew$LLOY)
lsenew$ABF.log  <- log10(lsenew$ABF  + 1 - min(lsenew$ABF))
```

## Variable selection: leaps-and-bounds

```{r leaps, fig.cap="Adjusted R-squared vs. model size from leaps-and-bounds. Larger models fit better, with diminishing returns past ~15 predictors."}
leaps1 <- leaps(lsenew[, -c(2, 8, 13, 15)], lsenew$VOD,
                names = names(lsenew)[-c(2, 8, 13, 15)],
                method = "adjr2", nbest = 5)
leaps_res <- data.frame(size = leaps1$size, adjr2 = leaps1$adjr2, leaps1$which)
ggplot(leaps_res, aes(size, adjr2)) +
  geom_point(alpha = 0.5) +
  labs(x = "Model size", y = expression(paste("Adjusted ", R^2))) +
  theme_minimal()
```

A 13-predictor leaps model captures most of the signal:

```{r leaps-fit}
leaps14 <- lm(VOD ~ STJ + SPX + AUTO + ABDN + RMV + BATS + SMIN + ANTO + BA +
                PRU + CPG + SSE.log + ABF.log, data = lsenew)
leaps14_r2 <- summary(leaps14)$adj.r.squared
```

This model reaches an adjusted $R^2$ of `r round(leaps14_r2, 4)`. The
diagnostic plots below suggest the residuals are not yet linear in the
fitted values — a Box-Cox response transform should help.

```{r leaps-diag, fig.height=4.5, fig.cap="Standard lm diagnostic plots for the leaps model. The QQ-plot shows heavier tails than ideal."}
par(mfrow = c(2, 2), mar = c(4.5, 4, 2, 2))
plot(leaps14)
```

```{r leaps-boxcox, fig.cap="Box-Cox profile for the leaps model. The peak near lambda ~= 2.6 suggests a power transform on VOD.", fig.height=4}
boxcox(leaps14, plotit = TRUE, lambda = seq(-4, 4))
```

Applying $VOD^{2.6}$ and refitting:

```{r leaps-transformed}
lsenew$VOD_train1 <- lsenew$VOD^2.6
leaps14.2 <- lm(VOD_train1 ~ TSCO + SPX + AUTO + ABDN + RMV + BATS + SMIN +
                  ANTO + BA + PRU + CPG + SSE.log + ABF.log, data = lsenew)
leaps14.2_r2 <- summary(leaps14.2)$adj.r.squared
```

The transformed model reaches adjusted $R^2 = `r round(leaps14.2_r2, 4)`$.
The QQ-plot below is noticeably straighter, though the residuals-vs-fitted
plot still has some structure.

```{r leaps-transformed-diag, fig.height=4.5, fig.cap="Diagnostic plots for the Box-Cox-transformed leaps model."}
par(mfrow = c(2, 2), mar = c(4.5, 4, 2, 2))
plot(leaps14.2)
```

## Variable selection: stepwise / forward / backward

We apply the three classical step-selection strategies to the same candidate
predictor set, with a Box-Cox transform on the response at $\lambda = 2.8$
(based on the boxcox output above). The original draft had two bugs that
are fixed here:

* **Self-referential `data` argument.** The backward fit referenced
  `data = lsenew4` inside the `subset()` call that *defined* `lsenew4`. It
  worked by lazy evaluation but was fragile. The data is now built first.
* **Malformed residual-inspection loop.** `for (i in x){ x <- c(...) }`
  reassigned the iteration variable inside its own body. Replaced with a
  simple `for (v in candidate_vars)` loop.

```{r selection}
candidate_vars <- c("Month", "STJ", "MGGT", "TSCO", "SPX", "AUTO", "AHT", "RTO",
                    "ABDN", "RMV", "SDR", "BATS", "ENT", "RR", "SMIN", "ANTO",
                    "BA", "PSN", "PRU", "CPG", "WTB",
                    "SSE.log", "LLOY.log", "ABF.log")
scope_formula <- reformulate(candidate_vars, response = "VOD")

# Drop the raw versions of the transformed predictors so we don't fit both.
lsenew_sel <- subset(lsenew, select = -c(SSE, LLOY, ABF))

# Stepwise (both)
stepwise_start <- lm(VOD ~ 1, data = lsenew_sel)
stepwise_picked <- step(stepwise_start, scope = scope_formula,
                        direction = "both", trace = 0)

# Forward
forward_start <- lm(VOD ~ 1, data = lsenew_sel)
forward_picked <- step(forward_start, scope = scope_formula,
                       direction = "forward", trace = 0)

# Backward: build the data frame FIRST (bug fix), then fit the full model
# and step down.
backward_data <- lsenew_sel[, c("VOD", candidate_vars)]
backward_data <- backward_data[complete.cases(backward_data), ]
backward_full <- lm(VOD ~ ., data = backward_data)
backward_picked <- step(backward_full, scope = scope_formula,
                        direction = "backward", trace = 0)

# Box-Cox-transformed response, same selection process.
lsenew_sel$VOD.three <- lsenew_sel$VOD^2.8
scope_formula_t <- reformulate(candidate_vars, response = "VOD.three")

stepwise_t <- step(lm(VOD.three ~ 1, data = lsenew_sel),
                   scope = scope_formula_t, direction = "both", trace = 0)
forward_t  <- step(lm(VOD.three ~ 1, data = lsenew_sel),
                   scope = scope_formula_t, direction = "forward", trace = 0)
backward_data_t <- lsenew_sel[, c("VOD.three", candidate_vars)]
backward_data_t <- backward_data_t[complete.cases(backward_data_t), ]
backward_t <- step(lm(VOD.three ~ ., data = backward_data_t),
                   scope = scope_formula_t, direction = "backward", trace = 0)
```

```{r selection-table}
selection_results <- data.frame(
  Method = c("Leaps (raw)", "Stepwise (raw)", "Forward (raw)", "Backward (raw)",
             "Leaps (transformed)", "Stepwise (transformed)",
             "Forward (transformed)", "Backward (transformed)"),
  Predictors = c(length(attr(terms(leaps14), "term.labels")),
                 length(attr(terms(stepwise_picked), "term.labels")),
                 length(attr(terms(forward_picked), "term.labels")),
                 length(attr(terms(backward_picked), "term.labels")),
                 length(attr(terms(leaps14.2), "term.labels")),
                 length(attr(terms(stepwise_t), "term.labels")),
                 length(attr(terms(forward_t), "term.labels")),
                 length(attr(terms(backward_t), "term.labels"))),
  Adj_R2 = c(summary(leaps14)$adj.r.squared,
             summary(stepwise_picked)$adj.r.squared,
             summary(forward_picked)$adj.r.squared,
             summary(backward_picked)$adj.r.squared,
             summary(leaps14.2)$adj.r.squared,
             summary(stepwise_t)$adj.r.squared,
             summary(forward_t)$adj.r.squared,
             summary(backward_t)$adj.r.squared)
)
knitr::kable(selection_results, digits = 4,
             col.names = c("Method", "Predictors", "Adjusted $R^2$"),
             escape = FALSE,
             caption = "Comparison of selection methods")
```

All three step-selection methods converge to similar models with similar
adjusted $R^2$ values. The transformed-response models fit better in-sample,
but on this kind of data that partly reflects the response transform magnifying
the apparent signal rather than genuine added structure.

## Residual diagnostics and a key caveat

The `lm` diagnostic plots (not shown for every model — see the `.R` script for
the full gallery) are reasonable: the QQ-plot has heavier tails than ideal
but no dramatic outliers, and the scale-location plot is largely flat. The
one diagnostic that flags a real problem is the **residual ACF**:

```{r acf, fig.height=3.5, fig.cap="ACF of the backward-transformed model residuals. Significant lag-1 autocorrelation confirms the residuals are not independent."}
acf(resid(backward_t), main = "")
```

There is strong lag-1 autocorrelation in the residuals. This is exactly what
we should expect when fitting OLS to *price levels* — the standard errors are
underestimated and the t-statistics are overstated. The honest conclusion is
not "all these predictors are significant" but "this is an explanatory fit
that ignores the time-series structure of the data". A proper analysis would
either model returns (or log-returns) directly, or use an ARIMAX / dynamic
regression that accounts for residual autocorrelation.

## PRESS statistic

As a single leave-one-out cross-validation summary, the PRESS statistic for
the leaps model is:

```{r press}
press <- sum((leaps14.2$residuals^2) / (1 - hatvalues(leaps14.2))^2)
```

PRESS = `r format(press, scientific = FALSE, digits = 5)`. Like the in-sample
$R^2$, this should be read with the autocorrelation caveat in mind — it does
not protect against the systematic over-fitting that residual autocorrelation
implies.

## Discussion

* Four classical variable-selection strategies (leaps, stepwise, forward,
  backward) all find that a subset of roughly 13–20 of the original 27
  predictors captures almost all of the explainable variation in VOD.
* A Box-Cox response transform around $\lambda \in [2.6, 2.8]$ improves the
  residual QQ-plot but does not change which predictors are selected.
* The residuals are strongly autocorrelated, which is the expected consequence
  of fitting OLS to price levels. The in-sample fit numbers look impressive
  but the standard errors should not be taken at face value.
* On synthetic data the workflow runs end-to-end and produces results in the
  same qualitative range as the original 2023 analysis (adj-$R^2$ around
  0.91, similar predictor counts). It is reproducible; the original
  coursework numbers are not, because the original `lse.RData` is lost.

## Reproducibility

Refresh the data with `Rscript data/generate_demo_data.R`. Re-render with
`rmarkdown::render("report/stock-regression-analysis.Rmd")`.
